---
title: "Business Forecasting" #appear to be mandatory
site: bookdown::bookdown_site #appear to be mandatory
---
# setup {-}

All the output formatting is done in the _output.yml file while all bookdown rendering options is done in the _bookdown.yml file.ยง

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE
                      ,message = FALSE
                      ,error = FALSE
                      ,warning = FALSE
                      ,strip.white = FALSE #To remove white space before and after chunk in output
                      ,fig.align = 'center'
                      ,fig.width = 5
                      )
```


```{r}
library(bookdown)
```

**Tips and tricks**

*Referencing*

This is making a figure with a caption and where we center the figure and set the size. Notice, that the figure is automatically numbered according to the chapter number.

```{r nice-fig, fig.cap="A fig", out.width='80%', fig.asp=.75, fig.align='center'}
plot(pressure,type = 'b',pch = 19)
```

Now we are able to make a reference to the chunk.

Refer to a chunk: \@ref(fig:nice-fig)

*Options*

[RMarkdown Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf?_ga=2.47682403.59925266.1608407175-330761457.1608407175)

